Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-09 main.tex
revise proof prop. closure and mv def. effectiveness

\begin{proposition} \label{prop:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$,
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and a $\SWA$ $A$ over $\Sigma$, $\Semiring$ and $\bar\Phi$,
there exists an effectively constructible \SWA
$B_{A, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$,
$B_{A, T}(t) = \displaystyle\bigoplus_{s\in \Sigma^*} A(s) \otimes T(s, t)$.
\end{proposition}
%
\begin{proof}
\florent{proof: sketch and mv to app.?}
Let $T = \< Q, \init_T, \bar{\wei}, \final_T >$,
where $\bar{\wei}$ contains $\wei_{10}$, $\wei_{01}$, and $\wei_{11}$,
from $Q \times Q$ into respectively
$\Phi_{\Sigma}$, $\Phi_{\Delta}$, and $\Phi_{\Sigma, \Delta}$,
and let $A = \< P, \init_A, \wei_1, \final_A >$
with $\wei_1: Q \times Q \to \Phi_{\Sigma}$.
%
\noindent
The state set of $B_{A, T}$ will be $Q' = P \times Q$.
The entering, leaving and transition functions of $B_{A, T}$ will
simulate synchronized computations of $A$ and $T$,
while reading an output word of $\Delta^*$.
%
Its state entering functions is defined
for all $p \in P$, $q \in Q$
by $\init'(p, q) = \init_A(p) \otimes \init_T(q)$.
The transition function $\wei'_1$ will roughly perform
a synchronized product of transitions defined by $\wei_1$,
$\wei_{01}$ ($T$ reading in output word and not an input word)
and $\wei_{11}$ ($T$ reading both an input word and an output word).
%
Moreover, $\wei'_1$ also needs to simulate transitions
defined by $\wei_{10}$: $T$ reading in input word and not an output word.
Since $B_{A, T}$  will read only in the output word, such a transition corresponds
to an $\varepsilon$-transition of $\SWA$.
But $\SWA$ have been defined without $\varepsilon$-transitions.
Therefore, in order to take care of this case, we perform an on-the-fly
suppression of $\varepsilon$-transition in the $\SWA$ in construction,
following the algorithm of~\cite{LombardySakarovitch12ciaa}.
%
% Initialize state leaving functions i
%and $\final'(p, q) = \final_A(p) \otimes \final_T(q)$.

%Every transition of $B_{A, T}$ will
%simulate a sequence of transitions of $T$ performing the following steps:
%advance in the input word while staying immobile in the output word,
%and then make one step in the output word (and advance in the input word or not).

\noindent
Initially, for all $p_1, p_2 \in P$, and $q_1, q_2 \in Q$, let
\[
\begin{array}{rcl}
\wei'_1\bigl( \< p_1, q_1>, \< p_2, q_2>\bigr) & = &
\wei_1(p_1, p_2) \otimes
\bigl[
\wei_{01}(q_1, q_2)
\oplus
\displaystyle\bigoplus_{\Sigma}
\wei_{11}(q_1, q_2)
\bigr]\\
\final'(p_1, q_1) & = & \final_A(p_1) \otimes \final_T(q_1)
\end{array}
\]

\noindent
Then, we iterate the following for all $p_1\in P$ and $q_1, q_2 \in Q$:
for all $p_2\in P$ and $q_3 \in Q$,
\[
\begin{array}{rcl}
\wei'_1\bigl( \< p_1, q_1>, \< p_2, q_3>\bigr) & \opluseq &
\displaystyle\bigoplus_{\Sigma} \wei_{10}(q_1, q_2)
\otimes
\wei'_1\bigl( \< p_1, q_2>, \< p_2, q_3>\bigr)\\
\final'(p_1, q_1) & \opluseq &
\displaystyle\bigoplus_{\Sigma} \wei_{10}(q_1, q_2)
\otimes \final'(p_1, q_2)
\end{array}
\]
\florent{proof correctness (app)}
%\qed
\end{proof}

@@

Formally, the transitions of the automaton~$A$ are defined
in term of %a weight value computed by
an intermediate function $\weight_A$, like in Section~\ref{sec:SWT}.
%In the case of a pushdown automaton,
A configuration, denoted by $q[\gamma]$,
is here composed of a state $q \in Q$
and a stack content $\gamma \in \Gamma^*$,
where $\Gamma = \Deltac \times P$.
Hence, $\weight_A$ is a function from
$[Q \times \Gamma^*] \times \Delta^* \times [Q \times \Gamma^*]$ into~$\Semiring$.
The empty stack is denoted by $\bot$, and the first %upmost
symbol is the last pushed content.
%
The recursive definition of $\weight_A$
enumerates each of the six possible cases:
reading $a \in \Deltai$,
%\lydia{intro to func}
or $\call{c} \in \Deltac$, or $\return{r} \in \Deltar$,
for each possible state of the stack (empty or not).
%to add to $u \in {\Delta}^*$.
%\lydia{introduced the 6 cases}
\florent{shorter notation $cp$ for $\< c, p>$ ?}

\begin{align}
%\weight_A\bigl(\config{q}{\bot}, \varepsilon, \config{q}{\bot}) & = \one\label{eq:SWVPA-weight}\\
%\weight_A\bigl(\config{q}{\bot}, \varepsilon, \config{q'}{\gamma}) & = \zero
%\mathrm{~if~} q \neq q'\nonumber\\
%
\weight_A\bigl(\config{q}{\bot}, \varepsilon, \config{q'}{\bot}) & = \one
\mathrm{~if~} q = q' \mathrm{~and~} \zero \mathrm{~otherwise}\label{eq:SWVPA-weight}\\
\weight_A\bigl(\configup{q}{\<{\call{c}}, p>\stackup\gamma}, a\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{q'' \in Q}} \weii(q, c, p, a, q'')
  \otimes \weight_A\bigl(\configup{q''}{\<{\call{c}}, p> \stackup \gamma}, u,
                         \config{q'}{\gamma'}\bigr)\nonumber\\
%
\weight_A\bigl(\config{q}{\bot}, a\, u,
               \config{q'}{\gamma'}\bigr) & =
  {\displaystyle\bigoplus_{q'' \in Q}} \weiei(q, a, q'')
   \otimes \weight_A\bigl(\config{q''}{\bot}, u, \config{q'}{\gamma'}\bigr)\nonumber\\
%
\weight_A\bigl(\configup{q}{\<{\call{c}}, p>\stackup\gamma}, {\call{c}'}\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{\begin{array}{c}
                          \scriptstyle q'' \in Q\\[-1ex]
                          \scriptstyle p' \in P
                          \end{array}}}
 \weic\bigl(q, {\call{c}}, p, {\call{c}'}, p', q''\bigr)
 \otimes \weight_A\bigl(\configup{q''}{\<{\call{c}'}, p'>\stackup \<{\call{c}}, p>\stackup \gamma}, u,
                        \config{q'}{\gamma'}\bigr)\nonumber\\[1mm]
%
\weight_A\bigl(\config{q}{\bot}, {\call{c}}\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{\begin{array}{c}
                          \scriptstyle q'' \in Q\\[-1ex]
                          \scriptstyle p \in P
                          \end{array}}}
  \weiec(q, {\call{c}}, p, q'')
  \otimes \weight_A\bigl(\config{q''}{\<{\call{c}}, p>}, u,
                         \config{q'}{\gamma'}\bigr)\nonumber\\
%
\weight_A\bigl(\configup{q}{\<{\call{c}}, p>\stackup\gamma}, {\return{r}}\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{q'' \in Q}}
  \weir\bigl(q, {\call{c}}, p, {\return{r}}, q''\bigr)
  \otimes \weight_A\bigl(\config{q''}{\gamma}, u,
                         \config{q'}{\gamma'}\bigr)\nonumber\\
%
\weight_A\bigl(\config{q}{\bot}, {\return{r}}\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{q'' \in Q}} \weier(q, {\return{r}}, q'')
  \otimes \weight_A\bigl(\config{q''}{\bot}, u,
                         \config{q'}{\gamma'}\bigr)\nonumber
\end{align}
\lydia{c p to <c, p>}
%
% and ${\call{c}}\, p\stacksep \gamma$
%denotes a stack where the pair made of ${\call{c}} \in \Deltac$ and $p \in P$ is the top symbol
%and $\gamma$ is the rest of stack.

@@

%\subsection{Properties}
\medskip\noindent
Similarly to \VPA~\cite{AlurMadhusudan09nested}
and \SVPA~\cite{dAntonyAlur14SVPDA},
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}\label{prop:SWVPA-product}
Let $A_1$ and $A_2$ be two \SWVPA
over the same $\Delta$, $\Semiring$ and $\bar\Phi$.
There exists two effectively constructible $\SWVPA$
$A_1 \oplus A_2$ and $A_1 \otimes A_2$,
such that for all $s \in \Delta^*$,
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
%
\begin{proof}
The construction is similar
to that of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.
\florent{complete proof (app)}
\end{proof}
\noindent
We shall now present a procedure for searching, for a~$\SWVPA$ $A$,
a word %$w \in \Delta^*$
of minimal weight for~$A$,
as stated in the following proposition. %\wrt~$\leq_\oplus$,
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a Dijkstra algorithm computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
it holds that:
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
The infinite sum in~\eqref{eq:bbot} is well defined since $\Semiring$ is complete.
%
Following~\eqref{eq:SWVPA-value}, and the associativity and commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is:
\begin{equation}\label{eq:min}
{\displaystyle\bigoplus_{t\in \Delta^*}} A(t)
=
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\end{equation}

@@

Termination: the algorithm performs $2.|Q|^2$ iterations until $P$ is empty,
and each iteration has a time complexity $O(|Q|^2 . |P|)$.
\florent{complexity: detail with nb tr. and states?}
That gives a time complexity $O(|Q|^4 . |P|)$.
It can be reduced by implementing $P$ as a priority queue,
prioritized by the value returned by $d$.
%$|Q|^3.\log(|Q|^2)$
%
At termination, $d_\bot = b_\bot$ and $d_\top = b_\top$
(see Appendix~\ref{sec:bestsearch}).
With~\eqref{eq:min}, this ensures
the correctness of Algorithm~\ref{algo:Dijkstra}.

\noindent
** effectively computable by hypothesis that the label theory is effective**
\florent{REVOIR}
In order to obtain effectively a witness
(word of $\Delta^*$ with a computation of $A$ of minimal weight),
we require the additional property of convexity of weight functions.***

>>>>

\begin{proposition} \label{prop:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$,
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$, %effective,
and a $\SWA$ $A$ over $\Sigma$, $\Semiring$ and $\bar\Phi$,
there exists an effectively constructible \SWA
$B_{A, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$,
$B_{A, T}(t) = \displaystyle\bigoplus_{s\in \Sigma^*} A(s) \otimes T(s, t)$.
\end{proposition}
\florent{AV: not needed $\bar\Phi$ effective}
%
\begin{proof}
\florent{proof: sketch and mv to app.?}
\input{app-closure}
%\qed
\end{proof}

@@

Formally, the transitions of the automaton~$A$ are defined
with an intermediate function $\weight_A$, like in Section~\ref{sec:SWT}.
%
A configuration $q[\gamma]$
is composed of a state $q \in Q$
and a stack content $\gamma \in \Gamma^*$,
where $\Gamma = \Deltac \times P$.
Hence, $\weight_A$ is a function from
$[Q \times \Gamma^*] \times \Delta^* \times [Q \times \Gamma^*]$ into~$\Semiring$.
The empty stack is denoted by $\bot$, and the first %upmost
symbol is the last pushed content.
%
The recursive definition of $\weight_A$
enumerates each of the six possible cases:
reading $a \in \Deltai$,
%\lydia{intro to func}
or $\call{c} \in \Deltac$, or $\return{r} \in \Deltar$,
and for each possible state of the stack (empty or not).
%to add to $u \in {\Delta}^*$.
%\lydia{introduced the 6 cases}
\florent{shorter notation $cp$ for $\< c, p>$ ?}

\begin{align}
%\weight_A\bigl(\config{q}{\bot}, \varepsilon, \config{q}{\bot}) & = \one\label{eq:SWVPA-weight}\\
%\weight_A\bigl(\config{q}{\bot}, \varepsilon, \config{q'}{\gamma}) & = \zero
%\mathrm{~if~} q \neq q'\nonumber\\
%
\weight_A\bigl(\config{q}{\bot}, \varepsilon, \config{q'}{\bot}) & = \one
\mathrm{~if~} q = q' \mathrm{~and~} \zero \mathrm{~otherwise}\label{eq:SWVPA-weight}\\
\weight_A\bigl(\configup{q}{\<{\call{c}}, p>\stackup\gamma}, a\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{q'' \in Q}} \weii(q, c, p, a, q'')
  \otimes \weight_A\bigl(\configup{q''}{\<{\call{c}}, p> \stackup \gamma}, u,
                         \config{q'}{\gamma'}\bigr)\nonumber\\
%
\weight_A\bigl(\config{q}{\bot}, a\, u,
               \config{q'}{\gamma'}\bigr) & =
  {\displaystyle\bigoplus_{q'' \in Q}} \weiei(q, a, q'')
   \otimes \weight_A\bigl(\config{q''}{\bot}, u, \config{q'}{\gamma'}\bigr)\nonumber\\
%
\weight_A\bigl(\configup{q}{\<{\call{c}}, p>\stackup\gamma}, {\call{c}'}\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{\begin{array}{c}
                          \scriptstyle q'' \in Q\\[-1ex]
                          \scriptstyle p' \in P
                          \end{array}}}
 \weic\bigl(q, {\call{c}}, p, {\call{c}'}, p', q''\bigr)
 \otimes \weight_A\bigl(\configup{q''}{\<{\call{c}'}, p'>\stackup \<{\call{c}}, p>\stackup \gamma}, u,
                        \config{q'}{\gamma'}\bigr)\nonumber\\[1mm]
%
\weight_A\bigl(\config{q}{\bot}, {\call{c}}\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{\begin{array}{c}
                          \scriptstyle q'' \in Q\\[-1ex]
                          \scriptstyle p \in P
                          \end{array}}}
  \weiec(q, {\call{c}}, p, q'')
  \otimes \weight_A\bigl(\config{q''}{\<{\call{c}}, p>}, u,
                         \config{q'}{\gamma'}\bigr)\nonumber\\
%
\weight_A\bigl(\configup{q}{\<{\call{c}}, p>\stackup\gamma}, {\return{r}}\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{q'' \in Q}}
  \weir\bigl(q, {\call{c}}, p, {\return{r}}, q''\bigr)
  \otimes \weight_A\bigl(\config{q''}{\gamma}, u,
                         \config{q'}{\gamma'}\bigr)\nonumber\\
%
\weight_A\bigl(\config{q}{\bot}, {\return{r}}\, u,
               \config{q'}{\gamma'}\bigr) & =
 {\displaystyle\bigoplus_{q'' \in Q}} \weier(q, {\return{r}}, q'')
  \otimes \weight_A\bigl(\config{q''}{\bot}, u,
                         \config{q'}{\gamma'}\bigr)\nonumber
\end{align}
%\lydia{c p to <c, p>}
%
% and ${\call{c}}\, p\stacksep \gamma$
%denotes a stack where the pair made of ${\call{c}} \in \Deltac$ and $p \in P$ is the top symbol
%and $\gamma$ is the rest of stack.

@@

%\subsection{Properties}
\medskip\noindent
Similarly to \VPA~\cite{AlurMadhusudan09nested}
and \SVPA~\cite{dAntonyAlur14SVPDA},
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}\label{prop:SWVPA-product}
Let $A_1$ and $A_2$ be two \SWVPA
over the same $\Delta$, $\Semiring$ and $\bar\Phi$.
There exists two effectively constructible $\SWVPA$
$A_1 \oplus A_2$ and $A_1 \otimes A_2$,
such that for all $s \in \Delta^*$,
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
%
\begin{proof}
The construction is similar
to that of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.
\florent{complete proof (app)}
\end{proof}
\noindent
We shall now present a procedure for searching, for a~$\SWVPA$ $A$,
a word %$w \in \Delta^*$
of minimal weight for~$A$.
%as stated in the following proposition. %\wrt~$\leq_\oplus$,
It requires the following property of label theories.
%
\begin{definition}\label{def:effective}
A label theory is called \emph{effective} when
for all $\phi \in \Phi_\Sigma$ and $\eta \in \Phi_{\Sigma, \Delta}$,
$\bigoplus_{\Sigma} \phi$,
$\bigoplus_{\Delta}\bigoplus_{\Sigma} \eta$, and
$\bigoplus_{\Sigma}\bigoplus_{\Delta} \eta$
can be effectively computed from $\phi$ and $\eta$, 
and moreover, the number of symbols reaching these bounds is finite 
and can be effectively computed.
\end{definition}
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a Dijkstra algorithm computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
it holds that:
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
The infinite sum in~\eqref{eq:bbot} is well defined since $\Semiring$ is complete.
%
Following~\eqref{eq:SWVPA-value}, and the associativity and commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is:
\begin{equation}\label{eq:min}
{\displaystyle\bigoplus_{t\in \Delta^*}} A(t)
=
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\end{equation}

@@

Termination: the algorithm performs $2.|Q|^2$ iterations until $P$ is empty,
and each iteration has a time complexity $O(|Q|^2 . |P|)$.
\florent{complexity: detail with nb tr. and states?}
That gives a time complexity $O(|Q|^4 . |P|)$.
It can be reduced by implementing $P$ as a priority queue,
prioritized by the value returned by $d$.
** where  $T$ is the worst time complexity for an oracle computing the values
of Definition~\ref{def:effective}.
%$|Q|^3.\log(|Q|^2)$
%
At termination, $d_\bot = b_\bot$ and $d_\top = b_\top$
(see Appendix~\ref{sec:bestsearch}).
With~\eqref{eq:min}, this ensures
the correctness of Algorithm~\ref{algo:Dijkstra}.

\noindent
** effectively computable by hypothesis that the label theory is effective**
\florent{REVOIR}
In order to obtain effectively a witness
(word of $\Delta^*$ with a computation of $A$ of minimal weight),...
%we require the additional property of convexity of weight functions.***
using the second requirement in Definition~\ref{def:effective}.
